Leadership across cultures and contexts: an exploratory analysis

### 2020-05-15 leadership_across_cultures_contexts.Rmd
Minor edits

@henrich_evolution_2001 introduced this important distinction to social science disciplines outside of anthropology, agreeing with @tiger_dominance_1970 and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] on the evolutionary roots of dominance, but offering an evolutionary account of prestige rooted in social rather than sexual selection for culturally acquired competencies, which inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Our analyses support the cross-cultural importance of leader aggression and coercion, as well as culturally acquired competencies, but also indicate multiple features of leader qualities beyond dominance and prestige.

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader competencies and shamanism**. The individual competencies cluster (Figure \@ref(fig:cluster)) comprises variables associated with dominance-based leadership, including coercive authority and aggressiveness, as well as qualities associated with prestige-based leadership, including knowledgeable/intelligent and experienced/accomplished. These dimensions are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)).

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. Shamans often provide valued public goods, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. @winkelman1987cross found that in societies sampled from the HRAF ($n = 43$) shamans provided decision-making services within judiciary, economic, military, or political domains in 44% or more of societies, and held charismatic leadership roles in 19% of societies; @winkelman1992shamans emphasizes the influential positions of leadership shamans attain through their charisma, social unification, healing abilities, and use of supernatural powers to cause harm. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate superhuman abilities [@singh_cultural_2017].

@@

```{r fivefold, fig.width=10, fig.cap="The intersections of evidence for our five basic elements of leadership among all text records (an 'upset' plot) [@gehlenborg2019]. Horizontal bars (left): The total number of text records in each element. Dots (bottom): Elements that both have evidence (intersection). Vertical bars (top): The number of text records in that intersection. Red: The number of text records involving shamans or leaders with supernatural qualities. "}

@@

marketvars <-
  tibble(
    Element = c(
      'High status', 
      frmt('Cognitive capital', cognitive_vars),
      frmt('Somatic capital', physical_vars),
      frmt('Provide benefits', benefit_ability_vars),
      frmt('Impose costs', cost_ability_vars)
    ),
    `Component variables` = c(
    'High status',
    names(cognitive_vars),
    names(physical_vars),
    names(benefit_ability_vars),
    names(cost_ability_vars)
    )
  )
kable(marketvars, caption = 'Operationalization of our leadership components')
```

>>>>

@henrich_evolution_2001 introduced this important distinction to social science disciplines outside of anthropology, agreeing with @tiger_dominance_1970 and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] on the evolutionary roots of dominance, but offering an evolutionary account of prestige rooted in social rather than sexual selection for culturally acquired competencies. @henrich_evolution_2001 inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Our analyses support the cross-cultural importance of leader aggression and coercion, as well as culturally acquired competencies, but they also indicate that some important modifiications are required.

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader competencies and shamanism**. The individual competencies cluster (Figure \@ref(fig:cluster)A) comprises variables associated with dominance-based leadership, including coercive authority and aggressiveness, as well as qualities associated with prestige-based leadership, including knowledgeable/intelligent and experienced/accomplished. These dimensions are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)).

Shamanism appears to be a distinct form of leadership that combines a strategy of inducing fear, similar to the dominance strategy, but is based on knowledge and expertise, similar to the prestige strategy. The *Competencies* feature included supernatural and feared dimensions (Figure \@ref(fig:cluster)). Our *Shamanism* variable (which was not one of our original leadership dimensions) was predicted by words that evoke both dominance and prestige, such as *supernatural*, *power*, *fear*, and *knowledge* (Figure \@ref(fig:shamanism)A), and by four other leadership dimensions that also overlap with the dominance and prestige constructs: *Feared*, *Medicinal functions*, *Ritual functions*, and *Experience/accomplished*. @winkelman1987cross found that in societies sampled from the HRAF ($n = 43$) shamans provided decision-making services within judiciary, economic, military, or political domains in 44% or more of societies, and held charismatic leadership roles in 19% of societies; @winkelman1992shamans emphasizes the influential positions of leadership shamans attain through their charisma, social unification, healing abilities, and use of supernatural powers to cause harm. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate superhuman abilities [@singh_cultural_2017].

@@

```{r fivefold, fig.width=10, fig.cap="The intersections of evidence for our five basic elements of leadership among all text records (an 'upset' plot) [@gehlenborg2019]. Horizontal bars (left): The total number of text records in each element. Black dots: has evidence. Gray dots: does not have evidence. Vertical bars (top): The number of text records in that intersection. Red: The number of text records involving shamans or leaders with supernatural qualities. "}

@@

marketvars <-
  tibble(
    Element = c(
      frmt('Cognitive capital', cognitive_vars),
      frmt('Social capital', social_vars), 
      frmt('Somatic capital', physical_vars),
      frmt('Provide benefits', benefit_ability_vars),
      frmt('Impose costs', cost_ability_vars)
    ),
    `Component variables` = c(
    names(cognitive_vars),
    names(social_vars),
    names(physical_vars),
    names(benefit_ability_vars),
    names(cost_ability_vars)
    )
  )
kable(marketvars, caption = 'Operationalization of our leadership components')
```
